---
title: "RStudio Exercise 5"
author: "Pyry Sipilä"
date: "2 December 2018"
output: 
  html_document:
    toc: TRUE
---

#Chapter 5: Dimensionality reduction techniques

```{r, include=FALSE}
#read and wrangl the data
human <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human1.txt" , sep = "," , header = T)
#1
library(stringr)
str_replace(human$GNI, pattern=",", replace ="")
human <- mutate(human, GNI = as.numeric(GNI))

#2
keep <- c("Country", "Edu2.FM", "Labo.FM", "Edu.Exp", "Life.Exp", "GNI", "Mat.Mor", "Ado.Birth", "Parli.F")
human <- select(human,one_of(keep))

#3
# filter out all rows with NA values
human <- filter(human, complete.cases(human) == TRUE)

#4
human$Country

# define the last indice we want to keep
last <- nrow(human) - 7

# choose everything until the last 7 observations
human_ <- human[1:last, ]

# add countries as rownames
rownames(human_) <- human_$Country

#5

# remove the Country variable
human_ <- select(human_, -Country)
```

###1
```{r}
library(GGally)
ggpairs(human_)
  # calculate the correlation matrix,round it and print it
  cor_matrix<-cor(human_)
  cor_matrix <- cor_matrix %>% round(digits=2)
  cor_matrix
  # visualize the correlation matrix
  library(corrplot)
  corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos ="d",tl.cex=0.6)

```
Maternal mortality has an escpecially skewed distribution. Life expectation and expected education years are highly correlated and so are maternal mortality and adolescent births. Life expectancy and maternal mortality are highly negatively correlated.

###2 Principal componen analysis (unscaled)
```{r}
#unscaled data
pca_human <- prcomp(human_)

# create and print out a summary of pca_human
s <- summary(pca_human)
s

# rounded percetanges of variance captured by each PC
pca_pr <- round(100*s$importance[2, ], digits = 1)

# print out the percentages of variance
pca_pr

# create object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.5, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2], sub = "PC1 reflects maternal mortality and PC2 gross national income")
```


```{r}
human_std <- scale(human_)
pca_human_std <- prcomp(human_std)

# create and print out a summary of pca_human
s_std <- summary(pca_human_std)
s_std

# rounded percetanges of variance captured by each PC
pca_pr_std <- round(100*s_std$importance[2, ], digits = 1)

# print out the percentages of variance
pca_pr_std

# create object pc_lab to be used as axis labels
pc_lab_std <- paste0(names(pca_pr_std), " (", pca_pr_std, "%)")

# draw a biplot
biplot(pca_human_std, cex = c(0.5, 1), col = c("grey40", "deeppink2"), xlab = pc_lab_std[1], ylab = pc_lab_std[2], sub = "PC1 reflexts maternal mortality and adolescent births and PC2 women in labor force and parliament")
```
With standardized variables, the first principal components capture much less variance. That it, because all variables have same variance after standardization, which prevents variables with high variance dominating the principal componen analysis.For the same reason, with standarization PC2 reflects women in labor force and parliament but without standarizaton it recflects gross national income.

###4
In the PCA with standardized data, PC1 reflexts maternal mortality and adolescent births and PC2 women in labor force and parliament.

###5
```{r}
library(FactoMineR)
tea <- read.csv("http://factominer.free.fr/course/doc/data_MCA_Tea.csv",sep = ";", header = T)
head(tea)
str(tea)
dim(tea)
summary(tea)

plot <- gather(tea) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") 
plot + geom_bar() 

# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset
plot <- gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") 
plot + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

# multiple correspondence analysis
mca <- MCA(tea_time)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")

```
Dim 1 seems to capture those who like tea shops and unpacked tea. Probably they are people who want to enjoy high quality tea in a good atmosphere. Dim 2 seems to capture those who put something else in their tea. I wonder what that may be.