---
title: "R Markdown, RStudio Exercise 2"
author: "Pyry Sipilä"
output:
  html_document:
    toc: TRUE
---

# Chapter 2: Linear regression

*Describe the work you have done this week and summarize your learning.*

I have worked on data wrangling and linear regression. I think I have learned a lot. Please see below.

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  

#I will start by reading in the learning2014 data from my own data folder.

```{r}
learning2014 <- read.csv("//atkk/home/p/pyrysipi/Documents/Tutkimus/Kurssit/Introduction to Open Data Science 2018/Data/learning2014.csv")
```

The data were collected from students in Jyväskylä University in 2014-2015. The aim was to assess how attitude and different learning approaches predict success on an introductory statistics course.

More information on the data can be found here: <http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS2-meta.txt>

Below I will explore the structure and dimensions of the data.
```{r}
#dimensions of the data:
dim(learning2014)
#the data have 166 observations and 7 variables

#structure of the data
str(learning2014)

#First six rows of the data:
head(learning2014)
```

Next, I will describe the data with nice plots and summaries of the variables
```{r}
# A scatter plot matrix
pairs(learning2014[-1], col = learning2014$gender)
library(ggplot2)
install.packages("GGally")
library(GGally)
p1 <- ggpairs(learning2014, lower = list(combo = wrap("facethist", bins = 20)))
p1
summary(learning2014)
```
From the plots and summaries above, it can be seen that the data consists of 110 women and 56 men. Mean age is 25.5 years. Age is unevenly distributed so that there are some people who are much older than average but no people who are much younger than average. This is called a skewed distribution. Other variables have a distribution that at least resembles a normal distribution.

Attitude has a moderate correlation with points (r~0.4). (Better attitude is associated with better success on the course.) Deep (deep approach) has a moderate negative correlation (r~-0.3) with surf (surface approach). This adds to the validity of these constructs, because one would assume deep and surface approaches to be antagonistic with each other. Correlation between other variables are low (|r|<0.2)

Next, I will fit a linear regression model with points as the dependent variable.
```{r}
lm1 <- lm(Points ~ attitude + deep + stra, data = learning2014)
summary(lm1)
```

The intercept is statistically significant with an aplha level of 0.05 (p<0.05). Attitude is a statistically significant predictor of points but deep and stra are not. I will drop deep and stra from the model.
```{r}
lm2 <- lm(Points ~ attitude, data = learning2014)
summary(lm2)
```
For every one points increase in attitude, the student will gain on average ~3.5 points. R-squared is ~0.19, which means that the model explains 19% of the variation in Points. Because attitude is the only vairable in the model, it explains those 19% of variation in Points.

Next, I will draw the diagnostic plots of the linear model I have just constructed (lm2).
```{r}
plot(lm2, which = c(1,2,5))
```
The residuals vs fitted plot does not show any obvious patterns. This indicates that the model fits the data reasonably well and the data are homoscedastic.

In the normal Q-Q plot the dots are reasonably close to the diagonal line. This means that the distribution of the residuals is reasonably close to normal distributions. Therefore, the assumption of linear regression of normally distributed residuals is not violated. 

The residuals vs leverage plot does not identify any points with a high Cook's distance (no points are below the dashed read line indicating a Cook's distance of 0.5). This means that there are no outliers that would have potential to distort the regression model.
